Sequential rank agreement methods for comparison of lists

### 2015-08-11 latex/paper.Rnw
Finilizing pt 1

\begin{abstract}
  The comparison of alternative rankings of a set of items is a
  general and prominent task in applied statistics. Predictor
  variables are ranked according to magnitude of association with an
  outcome, prediction models rank subjects according to the
  personalized risk of an event, and genetic studies rank genes
  according to their difference in gene expression levels. This
  article constructs measures of the agreement of two or more ordered
  lists. We use the standard deviation of the ranks to define a
  measure of agreement that both provides an intuitive interpretation
  and can be applied to any number of lists even if some or all are
  incomplete or censored. The main idea is then a permutation based
  and graphically approach to assess the sequential changes of
  agreement as a function of the depth of the lists. This may help to
  identify change-points in the agreement of the lists. The
  usefulness of these tools are illustrated using gene rankings, and
  using data from two Danish ovarian cancer studies on where we assess
  the within and between agreement of different statistical classification
  methods.
\end{abstract}

@@

A common task of scientific research is to rank items obtained by some
form of data analysis. Ranked lists occur naturally in applications of
statistics. Regression methods rank predictor variables according to
magnitude of their association with an outcome, prediction models rank
subjects according to their risk of an event, and genetic studies rank
genes according to their difference in gene expression levels across
samples.

When several rankings of the same items are available, a common
research question is to what extent do they agree. Once agreement is
established a further question is regarding the depths of agreement
starting from the top of the ranked lists. A typical situation arises
in high-dimensional genomics studies when several analysis methods are
applied to rank a list of genes according to their association with a
phenotype, treatment effect or other outcome. A measure of agreement
of gene rankings obtained by different methods could often help to
identify which genes are worth to pursue in further experiments.

Several approaches exist to measure the ``distance'' between two
ranked lists. Of these, Kendall's $\tau$ \citep{kendall1948} and
Spearman’s footrule $\rho$ \citep{Spearman1910} are among the most
well-known, but these measures do not distinguish between agreement in
the top versus towards the bottom of the lists. \citet{Shieh1998}
proposed a weighted version of Kendall's $\tau$ where each pair of
rankings can be assigned different weights, and \citet{Yilmaz2008}
proposed the $\tau_{ap}$ which places higher emphasis on the top of
the lists.
%
Spearman’s footrule uses the ranks of the variables for calculation of
the distance and the use of ranks are also employed in the $M$ measure
of \citet{Bar-Ilan2006} where the reciprocal rank differences are used
to calculate the agreement measure.

More recent approaches consider the intersection of lists as the basis
for a similarity measure. However, simple intersection also places
equal weights on all depths of the list. Thus, \citet{Fagin2003} and
\citet{Webber2010} proposed modified intersections which put more
emphasis on the top of the lists. One proposal is denoted the average
overlap (AO) and used for comparison later on. Specifically,
\citet{Webber2010} define their rank-biased overlap (RBO) by weighting
with a converging series to ensure that the top is weighted higher
than the potentially non-informative bottom of the lists.

It is possible to use the existing methods to calculate agreement of
lists until a given depth, i.e., limited to the $d$ items of each
list. However, the interpretation may not be straightforward,
especially in the case of more than two lists.

In this article we introduce sequential rank agreement for measuring
agreement among ranked lists. We provide graphical tools to analyse
the depth of agreement of a set of ranked lists. The general idea is
to define agreement based on the sequence of ranks from the first $d$
elements in each list. As agreement metric we adapt the limits of
agreement known from agreement between quantitative variables
\citep{Carstensen2010}. Our proposed approach allows to compare more
than two lists simultaneously, provides a dynamic measure of agreement
as a function of the depth in the lists, places high weight on the top
of the list, accommodates censored/incomplete lists of varying
lengths, and has a natural interpretation that directly relates to the
ranks. The graphical illustration of sequential rank agreement
potentially allows to infer a change-point, i.e., a list depth where
there is a change in the agreement of the lists.

In settings where rankings are obtained from analysis of correlated
variables it is then possible to compare the actual results to the
expected agreement in non-informative data obtained by
randomization. In this sense our approach is a combination and
generalization of some of the ideas of \citet{Carterette2009} and
\citet{Boulesteix2009}. The former compares two rankings based on the
distance between them as measured by a multivariate Gaussian
distribution and the latter presents an overview of approaches for
aggregation of ranked lists including bootstrap and leave-one-out
jackknife approaches.

The manuscript is organized as follows: In the next section we define
sequential rank agreement for multiple ranked lists and discuss how to
handle incomplete/censored lists. In section 3 we present and discuss
approaches to evaluate the sequential rank agreement obtained from an
experiment and to identify the maximal depth until which the ranked
lists agree to certain extent. Finally we present three different
applications of the proposed sequential rank agreement approach to
real data before we discuss the findings along with possible
extensions. The approaches presented in this manuscript are available
in the R package \texttt{SuperRanker}.

@@

Consider a set of $P$ different items $X=\{X_1,\dots,X_P\}$ and a
ranking function $R: \{X_1,\dots,X_P\}\to \{1,\dots,P\}$, such that
$R(X_p)$ is the rank of item $X_p$. The inverse mapping $R^{-1}$
assigns to rank $r\in\{1,\dots,P\}$ the item $R^{-1}(r)$ that was
given rank $r$. An ordered list is the realization of a ranking
function $R$ applied to a data sample $X$ and we let
$R_1(X),\dots,R_L(X)$, $L\geq2$, denote ordered lists obtained from
$L$ different ranking functions applied to the same data $X$. We will
use the same notation for the equivalent scenario where we have
ordered lists obtained from applying the same ranking function to $L$
samples from the same population, i.e., for ranked lists
$R(X^1), \ldots, R(X^L)$.  Panels (a) and (b) of
Table~\ref{tab:example} show a schematic example of these
mappings. Thus if $R^{-1}(1)=X_{34}$ then item $X_{34}$ is ranked
first in list $l$, and similarly $R_l(X_{34})=1$.

@@

The agreement of the lists regarding the rank given to an item $X_p$ can be
measured by
\begin{equation}
  A(X_p) = f(R_1(X_p), \ldots, R_L(X_p)),
\end{equation}
for a distance function $f.$ Throughout this paper we will use the
sample standard error as our function $f$ and hence use
$$A(X_p) = \sqrt{\frac{\sum_{i=1}^L (R_i(X_p) - \bar{R}(X_p))^2}{L-1}},
$$
where $\bar{R}(X_p)$ is the average rank assigned to item $X_p$, but
other choices could be made. The sample standard error has an
interpretation as the average distance of the individual rankings of
the lists from the average ranking.

We now describe what is exemplified in Panel (c) of Table
\ref{tab:example} and how it can be used to define \emph{sequential
  rank agreement}. For an integer $1\le d\le P$ we define the unique
set of items found by merging the first $d$ elements of each of the
$L$ lists, i.e., the set of items ranked less than or equal to $d$ in
any of the lists:
\begin{equation}
S_d = \{R_l^{-1}(r) ; r \leq d, l = 1, \ldots, L \}.
\end{equation}
The \emph{sequential rank agreement} is the pooled standard deviation
of the items found in the set $S_d$:
\begin{equation}
\textrm{SRA}(d)= \sqrt{\frac{\sum_{\{p \in
      S_d\}}(L-1)A(X_p)^2}{(L-1)|S_d|}}, \label{def:sra}
\end{equation}
where $|S_d|$ refers to the cardinality of the set $S_d$. Values of
\textrm{SRA} close to zero suggests that the lists agree while larger
values suggests disagreement. If the ranked lists are identical then
the value of SRA will be zero for all depths $d$. The sequential rank
agreement can be interpreted as the average distance of the individual
rankings of the lists from the average ranking for each of the items
we have seen until depth $d$. Note that the terms that are part of
\eqref{def:sra} are not independent since each list contains the ranks
from 1 to $P$ exactly once. Hence we simply use the pooled standard
deviation as a measure of the average distance between rankings.

@@

With fully observed lists we can plot the sequential rank agreement
\eqref{def:sra} as a function of depth $d$. An example is seen in top
panels of Figure~\ref{fig:example1} where four analysis methods were
used to rank 3051 gene expression values measured on 38 tumor mRNA
samples in order to classify between acute lymphoblastic leukemia
(ALL) and acute myeloid leukemia (AML)
\citep{Golub1999}. Preprocessing of the gene expression data was done
as described in \citet{Dudoit2002} and the four different analysis
approaches were: marginal two-sample $t$ tests, marginal logistic
regression analyses, elastic net logistic regression
\citep{friedman2010regularization}, and marginal maximum information
content correlations (MIC) \citep{Reshef2011}. For the first two
methods, the genes were ranked according to $p$ value, for logistic
regression the genes were ordered by size of the corresponding
coefficients (after standardization), and MIC was ordered by
correlation which resulted in the top rankings seen in Table~\ref{tab1}.

The sequential rank agreement plotted in Figure~\ref{fig:example1}
shows how many ranks apart we on average expect the genes found in the
top $d$ part of the lists to be for each depth $d$.  It can be seen
that agreement is higher in the top of the lists (smaller values on
the $y$ axis) corresponding to \emph{better} agreement than in the
tails of the lists.

@@

@@

<<fig1a, fig=TRUE,echo=FALSE,include=FALSE,cache=TRUE,eval=TRUE>>=
ysize <- 1000
nreference <- 20
## Plot the line
plot(res[1:ysize], col="black", lwd=3, type="l", ylim=c(0,1300), ylab="Sequential rank agreement", xlab="Depth")

@@

\begin{figure}[tb]
   \begin{center}
 \includegraphics[width=.45\textwidth]{paper-fig1a}
 \includegraphics[width=.45\textwidth]{paper-fig2a}
 \includegraphics[width=.45\textwidth]{paper-fig1b}
 \includegraphics[width=.45\textwidth]{paper-fig2b}
 \end{center}
 \caption{Left panels: Sequential rank agreement for 4 different
   analysis methods applied to the 3051 genes in the Golub data. Right
   panels: Corresponding sequential rank agreement for the same data
   but where only the top 20 ranked items are available. The blue and
   red areas correspond to the independent and randomized reference
   hypothesis areas, respectively. The bottom plots are identical to
   the corresponding top plots but have been zoomed in on the first
   part of the $x$ axis.}
 \label{fig:example1}
 \end{figure}

@@

\subsection{Analysis of incomplete/censored lists}

Incomplete lists are also a common occurrence. They arise for example
in case of missing data (items), when comparing top $d$ list results
from publications, or when some methods only rank a subset of the
items. For example, penalized regression based on the Lasso provides a
sparse set of predictors that have non-zero coefficients. There is no
obvious ordering of the set of predictors whose coefficient has been
shrunk to zero. Incomplete lists also occur when the analyst decides
to censor all lists at depth $d$ such that the measurement of
agreement is restricted to the first $d$ elements of each list.

Sequential rank agreement can be generalized to incomplete/censored
lists in the following way. Let $\Lambda_P$ be a set of $P$ items and
$\Lambda_l\subset \Lambda_P$ the subset of $d_l$ items that occur in
list $l$.  The case where all lists are censored at depth $d$
corresponds to $d_1=\cdots=d_L=d$. For incomplete/censored lists the
rank function becomes
\begin{equation}
  \widetilde R_l(X_p) = \left\{\begin{array}{cl} \{R_l(X_p)\} & \text{ for } X_p\in \Lambda_l \\
      \{d_l+1,\dots,P\} & \text{ for } X_p \not\in \Lambda_l\end{array}\right.
\end{equation}
where we only know that the rank for the unobserved items in list $l$
must be larger than the largest rank observed in that list.

The agreement, $A(X_p)$, cannot be computed directly for all
predictors in the presence of censored lists because the exact rank
for some items will be unknown. Also, the rankings within a single
list are not independent since each rank must appear exactly once in
each list. Thus, we cannot simply assign the same number (e.g., the
mean of the unassigned ranks) to the censored items since that would
result in less variation of the ranks and hence less variation of the
agreement, and it would artificially introduce a (downward) bias of
agreement for items that are censored in multiple lists.

Instead we randomize the ranks $\{d_{l}+1,\dots,P\}$ to the items that
do not occur in list $\Lambda_l$. By randomizing the missing items of
each list once we attain one realization of the $L$ rankings of the
set $\Lambda_P$. By randomizing a large number of times we can compute
\eqref{def:sra} for each realization, and then compute the sequential
rank agreement as the pointwise (for each depth) average of the rank
agreements. The algorithm is described in detail in
Algorithm~\ref{sra-algorithm}.

\begin{algorithm}
\caption{Sequential rank agreement algorithm for censored lists}
\label{sra-algorithm}
\begin{algorithmic}[1]
  \Procedure{Censored rank agreement}{} \State Let $B$ be the number
  of permutations to use \For{each $b \in B$} \For{each censored list
    $l \in L$}
  \State \parbox[t]{\dimexpr\linewidth-\algorithmicindent-1.7cm}{Permute
    the unassigned ranks, $\{d_l+1, \ldots, P\}$, and assign them
    randomly to the items \emph{not} found in the list, i.e.,
    $\Lambda^\complement=\Lambda_P\setminus\Lambda_l$. Combine the
    result with $\Lambda_l$ to fill out the list.}
\EndFor
\State \parbox[t]{\dimexpr\linewidth-\algorithmicindent-1cm}{Let sra($b$) be the sequential rank agreement computed from the
filled out lists.}
\EndFor
\State Return element-wise averages across all $B$ permutations of sra($b$).
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

\section{No information agreement}

@@

For the analysis of the Golub data described in Section
\ref{sec:amfol}, Figure~\ref{fig:example1} shows the empirical
distributions of sequential rank agreement under $H_0$ and
$\wildetilde H_0$ each based on $400$ permutations. Not surprisingly, the
sequential rank agreement under $\widetilde H_0$ is lower than the
sequential rank agreement under $H_0$ because the four methods used to
rank the data ($t$ test, logistic regression, elastic net, and MIC)
generally tend to identify (and rank high) similar predictors even if
there are only spurious associations. The two bottom panels in
Figure~\ref{fig:example1} also indicate that the observed sequential
rank agreement (the black line) is better than what would be expected
by chance for data that contain no information since it lies below the
reference areas. However, the censored data also suggests that there
may be at most 1 or 2 ranked items towards the top of the lists that
yield a result better than what would be expected (the bottom-right
plot).

>>>>

\begin{abstract}
  The comparison of alternative rankings of a set of items is a
  general and prominent task in applied statistics. Predictor
  variables are ranked according to magnitude of association with an
  outcome, prediction models rank subjects according to the
  personalized risk of an event, and genetic studies rank genes
  according to their difference in gene expression levels. This
  article constructs measures of the agreement of two or more ordered
  lists. We use the standard deviation of the ranks to define a
  measure of agreement that both provides an intuitive interpretation
  and can be applied to any number of lists even if some or all are
  incomplete or censored. The approach can identify change-points in
  the agreement of the lists and the sequential changes of agreement
  as a function of the depth of the lists can be compared graphically
  to a permutation based reference set.  The usefulness of these tools
  are illustrated using gene rankings, and using data from two Danish
  ovarian cancer studies where we assess the within and between
  agreement of different statistical classification methods.
\end{abstract}

@@

Ranking of items or results is common in scientific research and
ranked lists occur naturally as the result of many statistical
applications. Regression methods rank predictor variables according to
magnitude of their association with an outcome, prediction models rank
subjects according to their risk of an event, and genetic studies rank
genes according to their difference in gene expression levels across
samples.

When several rankings of the same items are available, a common
research question is to what extent they agree. In particular, is it
possible to identify an optimal rank until which the lists agree on
the items?  A typical situation arises in high-dimensional genomics
studies when several analysis methods are applied to rank a list of
genes according to their association with a phenotype, treatment
effect or other outcome. A measure of agreement of gene rankings
obtained by different methods could often help to identify which genes
are worth to pursue in further experiments.

Several approaches exist to measure the ``distance'' between two
ranked lists. Of these, Kendall's $\tau$ \citep{kendall1948} and
Spearman’s footrule $\rho$ \citep{Spearman1910} are among the most
well-known, but these measures do not distinguish between agreement in
the top versus towards the bottom of the lists and the provide a
single measure of the overall distance. \citet{Shieh1998} proposed a
weighted version of Kendall's $\tau$ where each pair of rankings can
be assigned different weights, and \citet{Yilmaz2008} proposed the
$\tau_{ap}$ which places higher emphasis on the top of the lists.
Spearman’s footrule uses the ranks of the variables for calculation of
the distance and the use of ranks are also employed in the $M$ measure
of \citet{Bar-Ilan2006} where the reciprocal rank differences are used
to calculate the similarity measure.

Other recent approaches consider the intersection of lists as the basis
for a similarity measure. However, simple intersection also places
equal weights on all depths of the list and therefore \citet{Fagin2003} and
\citet{Webber2010} proposed weighted intersections which put more
emphasis on the top of the lists. One proposal is denoted the average
overlap (AO) and used for comparison later on. Specifically,
\citet{Webber2010} define their rank-biased overlap (RBO) by weighting
with a converging series to ensure that the top is weighted higher
than the potentially non-informative bottom of the lists.
It is possible to use the existing methods to calculate agreement of
lists until a given depth, i.e., limited to the $d$ items of each
list. However, the interpretation may not be straightforward,
especially in the case of more than two lists, and they may not
accommodate partial rankings.

In this article we introduce sequential rank agreement for measuring
agreement among (partially) ranked lists.  The general idea is to
define agreement based on the sequence of ranks from the first $d$
elements in each list. As agreement metric we adapt the limits of
agreement known from agreement between quantitative variables
\citep{alt:bland:1983,Carstensen2010}. Our proposed approach allows us
to compare more than two lists simultaneously, it provides a dynamic
measure of agreement as a function of the depth in the lists, places
high weight on the top of the list, accommodates censored/incomplete
lists of varying lengths, and has a natural interpretation that
directly relates to the ranks. Graphical illustration of sequential
rank agreement potentially allows us to infer a change-point, i.e., a
list depth where a change in the agreement of the lists occur but we
also provide randomization-based graphical tools to compare the
observed rank agreement to the expected agreement found in
non-informative data.  In this sense our approach is a combination and
generalization of some of the ideas of \citet{Carterette2009} and
\citet{Boulesteix2009}. The former compares two rankings based on the
distance between them as measured by a multivariate Gaussian
distribution and the latter presents an overview of approaches for
aggregation of ranked lists including bootstrap and leave-one-out
jackknife approaches.

The manuscript is organized as follows: In the next section we define
sequential rank agreement for multiple ranked lists and discuss how to
handle incomplete/censored/partial lists. In section 3 we present and
discuss approaches to evaluate the results obtained from sequential
rank agreement. Finally we apply the sequential rank agreement to two
Danish ovarian cancer studies before we discuss the findings along
with possible extensions. The approaches presented in this manuscript
are available in the R package \texttt{SuperRanker}.

@@

Consider a set of $P$ different items $X=\{X_1,\dots,X_P\}$ and a
ranking function $R: \{X_1,\dots,X_P\}\to \{1,\dots,P\}$, such that
$R(X_p)$ is the rank of item $X_p$. The inverse mapping $R^{-1}$ gives
the item $R^{-1}(r)$ that was assigned to rank $r\in\{1,\dots,P\}$. An
ordered list is the realization of a ranking function $R$ applied to a
data sample $X$ and we let $R_1(X),\dots,R_L(X)$, $L\geq2$, denote
ordered lists obtained from $L$ different ranking functions applied to
the same data $X$. We will use the same notation for the equivalent
scenario where we have ordered lists obtained from applying the same
ranking function to $L$ samples from the same population, i.e., for
ranked lists $R(X^1), \ldots, R(X^L)$.  Panels (a) and (b) of
Table~\ref{tab:example} show a schematic example of these
mappings. Thus if $R_l^{-1}(1)=X_{34}$ then item $X_{34}$ is ranked
first in list $l$ and similarly $R_l(X_{34})=1$.

@@

The agreement of the lists regarding the rank given to an item $X_p$ can be
measured by
\begin{equation}
  A(X_p) = f(R_1(X_p), \ldots, R_L(X_p)),
\end{equation}
for a non-negative real-valued distance function $f.$ Throughout this
paper we will use the sample standard error as our function $f$ and
hence use
$$A(X_p) = \sqrt{\frac{\sum_{i=1}^L (R_i(X_p) - \bar{R}(X_p))^2}{L-1}},
$$
where $\bar{R}(X_p)$ is the average rank assigned to item $X_p$, but
other possibilities will be discussed below. For each item, the sample
standard error has an interpretation as the average distance of the
individual rankings from the average ranking over all the lists, and
it corresponds to the same measure that is used in method comparison
studies to compute the limits of agreement \citep{alt:bland:1983}.

For an integer $1\le d\le P$ we define the unique set of items found
by merging the first $d$ elements of each of the $L$ lists, i.e., the
set of items ranked less than or equal to $d$ in any of the lists:
\begin{equation}
S_d = \{R_l^{-1}(r) ; r \leq d, l = 1, \ldots, L \},
\end{equation}
which is exemplified in Panel (c) of Table
\ref{tab:example}.

The \emph{sequential rank agreement} is the pooled standard
deviation of the items found in the set $S_d$:
\begin{equation}
\textrm{sra}(d)= \sqrt{\frac{\sum_{\{p \in
      S_d\}}(L-1)A(X_p)^2}{(L-1)|S_d|}}, \label{def:sra}
\end{equation}
where $|S_d|$ refers to the cardinality of the set $S_d$. Values of
\textrm{sra} close to zero suggests that the lists agree while larger
values suggest disagreement. Just like in method comparison studies we
prefer smaller limits of agreement because that means that the
differences between the methods is small. If the ranked lists are
identical then the value of sequential rank agreement will be zero for
all depths $d$. The sequential rank agreement can be interpreted as
the average distance of the individual rankings of the lists from the
average ranking for each of the items we have seen until depth $d$.

Note that the terms that are part of \eqref{def:sra} are not
independent since every list contains the ranks from 1 to $P$ exactly
once. As a consequence we simply use the pooled standard deviation as
a measure of the average distance between rankings and we refrain from
using traditional distributional results for this measure.

@@

With fully observed lists we can plot the sequential rank agreement
\eqref{def:sra} as a function of depth $d$. An example is seen in top
panels of Figure~\ref{fig:example1} where four analysis methods were
used to rank 3051 gene expression values measured on 38 tumor mRNA
samples in order to classify between acute lymphoblastic leukemia
(ALL) and acute myeloid leukemia (AML)
\citep{Golub1999}. Preprocessing of the gene expression data was done
as described in \citet{Dudoit2002} and the four different analysis
approaches were: marginal two-sample $t$ tests, marginal logistic
regression analyses, elastic net logistic regression
\citep{friedman2010regularization}, and marginal maximum information
content correlations (MIC) \citep{Reshef2011}. For the first two
methods, the genes were ranked according to minimum $p$ value, for
logistic regression the genes were ordered by size of the
corresponding coefficients (after standardization), and MIC was
ordered by absolute correlation which resulted in the top rankings
seen in Table~\ref{tab1}.

The sequential rank agreement plotted in Figure~\ref{fig:example1}
shows how many ranks apart we on average expect the genes found in the
top $d$ part of the lists to be for each depth $d$.  It can be seen
that the sequential rank agreement limits are better towards the top
of the lists (smaller values on the $y$ axis) corresponding to
\emph{better} agreement than in the tails of the lists.

@@

options(SweaveHooks=list(fig=function() par(cex.lab=1.5, cex.axis=1.5, cex.main=1.5, cex.sub=1.5)))

@@

<<fig1a, fig=TRUE,echo=FALSE,include=FALSE,cache=TRUE,eval=TRUE>>=
ysize <- 1000
nreference <- 20
## Plot the line
plot(res[1:ysize], col="black", lwd=3, type="l", ylim=c(0,1300), ylab="Sequential rank agreement", xlab="Depth", cex=1.6)

@@

\begin{figure}[tb]
   \begin{center}
 \includegraphics[width=.45\textwidth]{paper-fig1a}
 \includegraphics[width=.45\textwidth]{paper-fig2a}
 \includegraphics[width=.45\textwidth]{paper-fig1b}
 \includegraphics[width=.45\textwidth]{paper-fig2b}
 \end{center}
 \caption{Left panels: Sequential rank agreement for 4 different
   analysis methods applied to the 3051 genes in the Golub data /black
   line). Right
   panels: Corresponding sequential rank agreement for the same data
   but where only the top 20 ranked items are available. The blue and
   red areas correspond to the independent and randomized reference
   hypothesis areas, respectively. The bottom plots are identical to
   the corresponding top plots but have been zoomed in on the first
   part of the $x$ axis.}
 \label{fig:example1}
 \end{figure}

@@

\subsection{Analysis of partial/censored lists}

Incomplete or partial lists are also a common occurrence. They arise
for example in case of missing data (items), when comparing top $d$
list results from publications, or when some methods only rank a
subset of the items. For example, penalized regression based on the
Lasso provides a sparse set of predictors that have non-zero
coefficients. There is no obvious ordering of the set of predictors
whose coefficient has been shrunk to zero and thus we end up with a
partial ordering. Incomplete lists also occur when the analyst decides
to censor all lists at depth $d$ such that the measurement of
agreement is restricted to the first $d$ elements of each list, or
when the analyst only wants to consider the ranking of items that have
been found to be statistically significant.

Sequential rank agreement can be generalized to partial lists in the
following way. Let $\Lambda_l\subset X$ be the subset of $d_l$ items
that have been ranked highest in list $l$.  The case where all lists are
censored at the same depth $d$ corresponds to $d_1=\cdots=d_L=d$. For
incomplete/censored lists the rank function becomes
\begin{equation}
  \widetilde R_l(X_p) = \left\{\begin{array}{cl} \{R_l(X_p)\} & \text{ for } X_p\in \Lambda_l \\
      \{d_l+1,\dots,P\} & \text{ for } X_p \not\in \Lambda_l\end{array}\right.
\end{equation}
where we only know that the rank for the unobserved items in list $l$
must be larger than the largest rank observed in that list.

The agreement, $A(X_p)$, cannot be computed directly for all
predictors in the presence of censored lists because the exact rank
for some items will be unknown. Also, recall that the rankings within
a single list are not independent since each rank must appear exactly
once in each list. Thus, we cannot simply assign the same number
(e.g., the mean of the unassigned ranks) to the censored items since
that would result in less variation of the ranks and hence less
variation of the agreement, and it would artificially introduce a
(downward) bias of agreement for items that are censored in multiple
lists.

Instead we randomize the ranks $\{d_{l}+1,\dots,P\}$ to the items that
do not occur in list $\Lambda_l$. One realization of the $L$ rankings
of the set $X$ is obtained by randomizing the missing items of each
list. By randomizing a large number of times we can compute
\eqref{def:sra} for each realization, and then compute the sequential
rank agreement as the pointwise (for each depth) average of the rank
agreements. The algorithm is described in detail in
Algorithm~\ref{sra-algorithm}.

\begin{algorithm}
\caption{Sequential rank agreement algorithm for censored lists}
\label{sra-algorithm}
\begin{algorithmic}[1]
  \Procedure{Censored rank agreement}{} \State Let $B$ be the number
  of permutations to use \For{each $b \in B$} \For{each censored list
    $l \in L$}
  \State \parbox[t]{\dimexpr\linewidth-\algorithmicindent-1.7cm}{Permute
    the unassigned ranks, $\{d_l+1, \ldots, P\}$, and assign them
    randomly to the items \emph{not} found in the list, i.e.,
    $\Lambda^\complement=X\setminus\Lambda_l$. Combine the
    result with $\Lambda_l$ to fill out the list.}
\EndFor
\State \parbox[t]{\dimexpr\linewidth-\algorithmicindent-1cm}{Let sra($b$) be the sequential rank agreement computed from the
filled out lists.}
\EndFor
\State Return element-wise averages across all $B$ permutations of sra($b$).
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

\section{Evaluating sequential rank agreement}

@@

For the analysis of the Golub data described in Section
\ref{sec:amfol}, Figure~\ref{fig:example1} shows the empirical
distributions of sequential rank agreement under $H_0$ and
$\widetilde H_0$ each based on $400$ permutations. Not surprisingly, the
sequential rank agreement under $\widetilde H_0$ is lower than the
sequential rank agreement under $H_0$ because the four methods used to
rank the data ($t$ test, logistic regression, elastic net, and MIC)
generally tend to identify (and rank high) similar predictors even if
there are only spurious associations. The two bottom panels in
Figure~\ref{fig:example1} also indicate that the observed sequential
rank agreement (the black line) is better than what would be expected
by chance for data that contain no information since it lies below the
reference areas. However, the censored data also suggests that there
may be at most 1 or 2 ranked items towards the top of the lists that
yield a result better than what would be expected (the bottom-right
plot).
